\pkg{partykit}: A Toolbox for Recursive Partytioning

### 2009-02-25 pkg/inst/slides/partykit_useR2009.tex
some re-phrasing and style touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@126 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[10pt]{article}

\usepackage{url}
\usepackage{natbib}

\renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
\newcommand{\affiliations}{\footnotesize}
\newcommand{\keywords}{\paragraph{Keywords:}}

\setlength{\oddsidemargin}{0cm} \setlength{\evensidemargin}{0cm}
\setlength{\textwidth}{16.5cm} \setlength{\topmargin}{-1cm}
\setlength{\textheight}{24.5cm}

@@

\title{\textbf{partykit}: A Toolbox for Recursive Partytioning}

\begin{center}
  {\bf Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}
\end{center}

\begin{affiliations}
1. Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen \par
2. Department of Statistics and Mathematics, Wirtschaftsuniversit\"at Wien \par
\texttt{\{Torsten.Hothorn, Achim.Zeileis\}@R-project.org}
\end{affiliations}

\keywords Regression Trees, Visualization, CTree, CHAID, Weka, PMML.

\vskip 0.8cm

Recursive Partitioning methods, or simply `trees', are simple yet powerful
methods to estimate regression relationships. Since the publication of the
Automated Interaction Detection (AID) algorithm in 1964, many extentions,
modifications and new approaches have been suggested in both the statistics
and machine learning communities. Most of the algorithms in use today are
available to the \textsf{R} user, for example from packages \textbf{rpart}
\citep{PKG:rpart},
\textbf{RWeka} \citep{PKG:RWeka}, \textbf{party} \citep{PKG:party}, or \textbf{mvpart}
\citep{PKG:mvpart}.

Computations on fitted trees or the implementation of a new or modified
recursive partitioning method, however, are relatively difficult tasks.
As an example, one might want to predict a numeric response by the median
instead of the mean based on an \textit{rpart} object. Or, instead of
binary splits, one is going to implement an algorithm which potentially
uses multiway splits, maybe containing a non-trivial model in each terminal
node. These, and other, tasks require extensive programming,
mainly because each package dealing with (some specialized form of)
recursive partitioning uses a different internal representation of a tree. Thus, no
standard user-interface exists, and methods for predictions, graphical or textual
visualizations, or other computations heavily rely on these internals.

To overcome these difficulties, the \textbf{partykit} package \citep{PKG:partykit} offers a 
unified representation of tree objects along with \texttt{predict}, 
\texttt{print}, and \texttt{plot} methods. Trees are represented through 
a class \textit{party}, with building blocks \textit{partynode} and 
\textit{partysplit}. The design of these classes covers splits in arbitrary
functions of (one or more) variables, including multiway splits. In
addition, non-constant fits in the terminal nodes are supported as well.
Conversion methods exist for the most important sources of recursive
partitioning objects, currently \textit{rpart}, \textit{J48} and
\textit{pmmlTreeModel} objects can easily be converted into a \textit{party}
object to compute with. The \textbf{partykit} also offers a re-implementation
of conditional inference trees \cite{Hothorn:2006:JCGS}, illustrating the usefulness of the generic
infrastructure.

In our presentation, we will only sketch details of these classes and 
corresponding methods and focus
on applications of the toolkit, for example for producing nice and
non-standard plots
for \textit{rpart} or \textit{J48} objects, for computing predictions
on millions of new observations really fast, and for implementing CHAID, 
a popular classification tree algorithm.

\bibliographystyle{plainnat}
\bibliography{partykit}

>>>>

\documentclass[10pt,a4paper]{article}

\usepackage{url,a4wide}
\usepackage[authoryear,round]{natbib}

\renewcommand{\title}[1]{\begin{center}{\bf \LARGE #1}\end{center}}
\renewcommand{\author}[1]{\begin{center}{\bf #1}\end{center}}
\newenvironment{affiliations}{\footnotesize \begin{enumerate}}{\end{enumerate}}
\newenvironment{keywords}{\paragraph{Keywords:}}{\bigskip}

\let\code=\texttt
\let\pkg=\textbf
\let\proglang=\textsf
\newcommand{\class}[1]{`\code{#1}'}

@@

\title{\pkg{partykit}: A Toolbox for Recursive Partytioning}
\author{Torsten Hothorn$^{1}$ and Achim Zeileis$^{2}$}

\begin{affiliations}
  \item Institut f\"ur Statistik, Ludwig-Maximilians-Universit\"at M\"unchen, Germany\\
        \texttt{Torsten.Hothorn@R-project.org}
  \item Department of Statistics and Mathematics, Wirtschaftsuniversit\"at Wien, Austria\\
        \texttt{Achim.Zeileis@R-project.org}
\end{affiliations}

\begin{keywords}
Regression Trees, Visualization, CTree, CHAID, Weka, PMML.
\end{keywords}

Recursive partitioning methods, or simply ``trees'', are simple yet powerful
methods for capturing regression relationships. Since the publication of the
automated interaction detection (AID) algorithm in 1964, many extentions,
modifications, and new approaches have been suggested in both the statistics
and machine learning communities. Most of the standard algorithms are
available to the \proglang{R} user, e.g., through packages
\pkg{rpart} \citep{PKG:rpart},
\pkg{RWeka} \citep{PKG:RWeka},
\pkg{party} \citep{PKG:party}, or
\pkg{mvpart} \citep{PKG:mvpart}.

However, no common infrastructure is available for representing trees
fitted by different packages. Consequently, the capabilities for extraction of
information---such as predictions, printed summaries, or visualizations---vary
between packages and come with somewhat different user interfaces.
Furthermore, extensions or modifications often require considerable
programming effort, e.g., if the median instead of the mean of a numerical
response should be predicted in each leaf of an \class{rpart} tree.
Similarly, implementations of new tree algorithms might also require new
infrastructure if they have features not available in the above-mentioned
packages, e.g., multi-way splits or more complex models in the leafs.

To overcome these difficulties, the \pkg{partykit} package \citep{PKG:partykit} offers a 
unified representation of tree objects along with \code{predict()}, 
\code{print()}, and \code{plot()} methods. Trees are represented through 
a new flexible class \class{party} which can, in principle, capture
all trees mentioned above but can also accomodate multi-way or functional
splits, as well as complex models in (leaf) nodes. The package is currently
under development at R-Forge but already provides conversion methods for
trees of classes \class{rpart}, \class{J48}, and \class{pmmlTreeModel}
as well as a re-implementation of conditional inference trees \citep{Hothorn:2006:JCGS}.

In our presentation, we will only sketch details of these classes and 
corresponding methods and focus on applications of the toolkit including
extended visualizations for \class{rpart} or \class{J48} objects, fast predictions
on millions of new observations, and a new implementation of the classical CHAID
algorithm.

\bibliographystyle{jss}
\bibliography{partykit}
